---
title: "Omnipath to hipathia"
author: "Martin Garrido Rodriguez-Cordoba"
output: github_document
---

## Abstract

The main goal of this parser is to transform an [Omnipath](http://omnipathdb.org/) formatted set of interactions into an object usable by the mechanistic modelling tool [hipathia](http://hipathia.babelomics.org/). Hipathia uses a signal propagation algorithm to estimate the activity of receptor-to-effector signaling circuits. Its current version uses a pathway-centric approach fed by a mixture of context-specific and generic signaling pathways as defined by the [KEGG  classification](https://www.genome.jp/kegg/pathway.html). On the other hand, Omnipath is a meta-resource that contains biological information from different databases in a ready-to-use tabular format. It comprises several types of relationships between biological entities as protein-protein interactions or TF-target relationships, as well as gene and protein functional annotations.

## Case of use

Load required packages

```{r, message = FALSE}
library(OmnipathR)
library(hipathia)
library(igraph)
library(dplyr)
library(Rgraphviz)
library(org.Hs.eg.db)
source("R/omnipath-to-hipathia.R")
```

Import omnipath interactions and subset to directed interactions which are not "auto" interactions.

```{r}
# import all interactions in omnipath and intercell information
interactions <- OmnipathR::import_Omnipath_Interactions()

# filter only to directed interactions (consensus) and remove self interactions
intInteractions <- subset(interactions, 
                          (consensus_stimulation == 1 & consensus_inhibition == 0) | 
                            (consensus_stimulation == 0 & consensus_inhibition == 1)) %>%
  subset(source_genesymbol != target_genesymbol)
# print head of interesting interactions
head(intInteractions) %>% OmnipathR::print_interactions()
```

To test the parser, we will create a network using functional annotations from KEGG. We will select the interactions between genes annotated within the "Retrograde endocannabinoid signaling" pathway.

```{r}
# import kegg annotations
keggAnnotations <- OmnipathR::import_Omnipath_annotations(filter_databases = "KEGG")
# subset to pathway of interest
intGenes <- subset(keggAnnotations, value == "Retrograde endocannabinoid signaling") %>% 
  pull(genesymbol)
selectedInteractions <- subset(intInteractions, source_genesymbol %in% intGenes & target_genesymbol %in% intGenes)
# print head of interesting interactions
head(selectedInteractions) %>% OmnipathR::print_interactions()
```

Use the **omnipathToHipathia** function to create the metaginfo object.  

```{r}
mgi <- omnipathToHipathia(selectedInteractions)
```

Plot the resulting graph object, which has been divided into circuits usable by the hipathia function.

```{r}
g <- mgi$pathigraphs$hsa00$graph
plot(g, layout = graphVizLayout(g), vertex.shape = "none")
```

## Note

This example is carried out with an extremely simple network. **The creation of the MGI object takes a really long time on complex graphs**, specially when there are lots of edges between nodes which are not receptors nor effectors.

## Session info

```{r}
sessionInfo()
```
