---
title: "Omnipath to hipathia"
author: "Martin Garrido Rodriguez-Cordoba"
date: "`r Sys.Date()`"
output: github_document
---

## Abstract

The main goal of this parser is to transform an [Omnipath](http://omnipathdb.org/) formatted set of interactions into an object usable by the mechanistic modelling tool [hipathia](http://hipathia.babelomics.org/). Hipathia uses a signal propagation algorithm to estimate the activity of receptor-to-effector signaling circuits. Its current version uses a pathway-centric approach fed by a mixture of context-specific and generic signaling pathways as defined by the [KEGG  classification](https://www.genome.jp/kegg/pathway.html). On the other hand, Omnipath is a meta-resource that contains biological information from different databases in a ready-to-use tabular format. It comprises several types of relationships between biological entities as protein-protein interactions or TF-target relationships, as well as gene and protein functional annotations.

The possible parser features derive from the curated information on Omnipath. Some examples:

1. Restrict to interactions from certain resources (i.e Reactome, BioGRID... etc).

2. Use the curation effort to create a range of results depending on the network reliability.

3. Expand the effector nodes to new effectors, using Kinase-Substrate or TF-target interactions.

4. Create a meta graph without filtering nodes by biological context.

**Disclaimer**: This repository includes the neccesary functions to perform the conversion of Omnipath interactions into hipathia MGI objects. **The creation of the MGI object takes a really long time on complex graphs**, specially when there are lots of edges between nodes which are not receptors nor effectors. 

## Packages and functions

Load required packages

```{r, message = FALSE}
# main packages
library(OmnipathR)
library(hipathia)
# annotation package to translate symbols into entrezs
library(org.Hs.eg.db)
# packages to handle and plot graphs
library(igraph)
library(ggplot2)
library(ggraph)
library(Rgraphviz)
library(cowplot)
library(ggVennDiagram)
# other packages from tidyverse
library(dplyr)
library(purrr)
library(tidyr)
library(tibble)
# custom parsing and plotting funs
source("R/omnipath-to-hipathia.R")
source("R/visualize-hipathia-graph.R")
# set ggplot default theme
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5)))
```

## Prepare Omnipath interactions 

Import all Omnipath interactions with `import_Omnipath_Interactions()`, which contains only interactions with references. Subset such interactions to consensus stimulation or inhibition and discard auto-interactions. The **curation_effort** attribute contains the the number of unique resource+reference pairs per interaction.

```{r}
interactions <- OmnipathR::import_Omnipath_Interactions()
# filter only to directed interactions (consensus) and remove self interactions
intInteractions <- subset(interactions, 
                          (consensus_stimulation == 1 & consensus_inhibition == 0) | 
                            (consensus_stimulation == 0 & consensus_inhibition == 1)) %>%
  subset(source_genesymbol != target_genesymbol)
```

Explore number of interactions per resource

```{r, warning = FALSE, fig.width=5, fig.height=9}
strsplit(intInteractions$sources, ";") %>%
  unlist() %>%
  data.frame(source = .) %>%
  count(source) %>%
  arrange(n) %>%
  mutate(source = factor(source, levels = source)) %>%
  ggplot(aes(x = n, y = source)) +
  geom_bar(stat = "identity")
```

For this case of use, we will restrict interactions to those included in Reactome and BioGRID

```{r}
selectedSources <- c("Reactome", "BioGRID")
intInteractions <- subset(intInteractions, grepl(pattern = paste0(selectedSources, collapse = "|"), x = sources))
```

## Hipathia default MGI

Hipathia uses a pathway-centric approach, which isolates nodes and interactions in different biological contexts defined by KEGG. It decomposes the signaling networks into a meta graph information object (MGI), which contains the main graph, the decomposed signaling circuits (in form of [igraph](https://igraph.org/) objects) and the neccesary metadata to carry out the analysis. For this case of use, we will focus on the [cell cycle pathway](https://www.genome.jp/kegg-bin/show_pathway?hsa04110).

```{r}
intPathways <- c("Cell Cycle"="hsa04110")
hipathiaMgi <- hipathia::load_pathways(species = "hsa", pathways_list = intPathways)
```

This is the complete network already processed to be used by Hipathia

```{r, fig.width=8, fig.height=5}
beautyHipathiaGraph(hipathiaMgi$pathigraphs$hsa04110$graph)
```

Some of the decomposed subgraphs (all receptors that reach one effector)...

```{r}
lapply(hipathiaMgi$pathigraphs$hsa04110$effector.subgraphs[1:2], beautyHipathiaGraph)
```

## Hipathia nodes, Omnipath interactions

For a first try, we will employ the subset of nodes from the [cell cycle pathway](https://www.genome.jp/kegg-bin/show_pathway?hsa04110), using Omnipath interactions to link the nodes. The first step consists on obtaining the list of genes (nodes) from hipathia:

```{r}
intGenes <- V(hipathiaMgi$pathigraphs$hsa04110$graph)$genesList %>%
  unlist() %>%
  mapIds(x = org.Hs.eg.db, keys = ., keytype = "ENTREZID", column = "SYMBOL") %>%
  as.character() %>%
  .[!is.na(.)]
```

And subset the Omnipath interactions to those which contain a relation between genes in the pathway **including complexes**.

```{r}
sourceIndex <- strsplit(intInteractions$source_genesymbol, "_") %>% 
  sapply(function(x) any(x %in% intGenes))
targetIndex <- strsplit(intInteractions$target_genesymbol, "_") %>% 
  sapply(function(x) any(x %in% intGenes))
# filter to interactions fulfilling the criteria
filteredInteractions <- intInteractions[sourceIndex & targetIndex, ]
```

Once with the selected interactions, we can apply the **omnipathToHipathia()** function to transform this network into the hipathia MGI object. This function relies on the hipathia function **mgi_from_sif()**, which creates the MGI object from the sif and attribute files. The **omnipathToHipathia()** will format the network and write the files into a temporary location in order to be imported with hipathia.

```{r, fig.width=8, fig.height=5}
omnipathMgi <- omnipathToHipathia(filteredInteractions)
```

We can now perform a visual comparison of the networks

```{r, fig.width=15, fig.height=7}
default <- beautyHipathiaGraph(g = hipathiaMgi$pathigraphs$hsa04110$graph) + ggtitle("Default Cell Cycle")
op <- beautyHipathiaGraph(g = omnipathMgi$pathigraphs$hsa00$graph) + ggtitle("Omnipath derived Cell Cycle")
cowplot::plot_grid(default, op)
```

And explore how many gene-to-gene interactions are shared between both MGIs

```{r, fig.width=10, fig.height=5}
# create a graph list
graphs <- list(default = hipathiaMgi$pathigraphs$hsa04110$graph, omnipath = omnipathMgi$pathigraphs$hsa00$graph)
# get all possible gene to gene interactions
geneToGene <- lapply(graphs, function(g) {
  dfs <- igraph::as_data_frame(g, what = "both")
  isNull <- sapply(dfs$vertices$genesList, is.null)
  dfs$vertices$genesList[isNull] <- NA
  geneToGene <- tibble(from = dfs$vertices[dfs$edges$from, "genesList"], to = dfs$vertices[dfs$edges$to, "genesList"]) %>%
    tidyr::unnest(from) %>%
    tidyr::unnest(to) %>%
    subset(from != "/" & !is.na(from) & to != "/" & !is.na(to)) %>%
    transmute(id = paste0(from, "-->", to)) %>% 
    pull(id)
  return(geneToGene)
})
# explore overlap between two sets
ggVennDiagram::ggVennDiagram(geneToGene)
```

The `omnipathMGI` object is now ready to be passed to functions from **hipathia** such as:

1. `hipathia()`
2. `normalize_paths()`
3. `get_path_names()`

## Session info

```{r}
sessionInfo()
```
